Mapping malaria by sharing spatial information between incidence and prevalence datasets

### 2019-09-04 ms.tex
KB edits.

Global malaria incidence has decreased dramatically over the last 20 years \citep{bhatt2015effect, weiss2019mapping, battle2019mapping}.
This decrease has been accompanied by a strategic shift aiming for elimination in low incidence countries \citep{world2016world, newby2016path}.
Accurate, high-resolution maps of malaria risk are vital in countries in the elimination and pre-elimination phases as they highlight the areas with ongoing \emph{Plasmodium} transmission most in need of interventions \citep{sturrock2016mapping, cohen2017mapping}.
Mapping malaria in low burden countries presents new challenges as traditional mapping of prevalence \citep{gething2011new, bhatt2017improved, gething2012long, bhatt2015effect} using cluster-level surveys and model-based geostatistics are not necessarily effective in these areas \citep{sturrock2016mapping, sturrock2014fine}.
In low burden areas, very large sample sizes are needed before a prevalence survey is informative because so few individuals have detectable parasitaemia that most sample points will have no cases.
In most cases, these large sample sizes are neither logistically nor financially feasible.
However, the availability and quality of routine surveillance data of malaria case counts, typically aggregated by administrative unit polygons, is improving, thus providing an alternative data source for mapping malaria burden \citep{sturrock2016mapping, ohrt2015information, cibulskis2011worldwide}.
Advantageously, the routine surveillance data can be more sensitive than prevalence point-surveys in low transmission areas because the entire public health system is being used to passively monitor disease occurrence \citep{cibulskis2011worldwide}.

Disaggregation regression methods have been proposed as a way to model malaria burden using polygon-level, routine surveillance records of incidence \citep{sturrock2014fine, wilson2017pointless, law2018variational, taylor2017continuous, li2012log}.
The intent of this research is to assess the utility of extending disaggregation regression modelling approaches to map malaria incidence using both point- and polygon-level response data. 
In particular it focuses on two aspects where improvements in the disaggregation regression may be possible:
(1) for modelling low-burden areas which have better coverage of prevalence surveys than polygon incidence data; and
(2) for building the statistical relationships between the response (i.e. malaria incidence) and geospatial predictor data in heterogeneous landscapes. 
An ancillary benefit of a hybridized approach is that it will simultaneously produce estimates of both prevalence and incidence metrics, which may both be useful for policy makers \citep{cohen2017mapping}.
Mdels are typically fitted to observations of one metric and then a secondary model is used to convert between prevalence and incidence \emph{post hoc} \citep{battle2019mapping, bhatt2015effect}, thus missing an opportunity to learn the relationship between prevalence and incidence at the same time as fitting the geographic model.

@@

We used two data sources that quantify malaria burden: prevalence point-surveys and polygon incidence data.
Prevalence point-surveys consist of geo-located survey clusters wherein all sampled individuals are tested for malaria and the positive cases as well as the total number of children tested is recorded. 
Polygon incidence data is aggregated to administrative units (e.g. districts or provinces) summarizing data reported from hospitals and health facilities. 
Unlike the point data, polygon-level reports only include numbers of cases and not the numbers of individuals in each administrative unit.  
As such, to determine an incidence rate we rely on gridded population surfaces, summarised to administrative unit boundaries, to provide the denominator. 
The prevalence point-survey data were extracted from the Malaria Atlas Project database \citep{bhatt2015effect, guerra2007assembling, pfeffer2018ma}. 
As the prevalence point-surveys cover different age ranges they were standardised to the 2--10 year-range using a previously published model \citep{smith2007standardizing}. 
As described, the age standardisation model gives the surveys with zero positive cases a small positive prevalence. 
The polygon incidence data were collated from various government reports and adjusted for incompleteness using methods defined by Cibulskis and colleagues \citep{cibulskis2011worldwide, weiss2019mapping}. 
These adjustments account for underreporting of clinical cases due to lack of treatment seeking, missing case reports (from a health facility that reported for 11 months in a year for example), and cases that sought medical attention outside the public health systems \citep{battle2016treatment}. 
Where species specific reports were given, these were used, and in reports that did not distinguish between species of \emph{Plasmodium} the national estimate of the ratio between \emph{P. falciparum} and \emph{P. vivax} cases was used to estimate numbers of \emph{P. falciparum} cases specifically. 
These adjustments were uniform across each country. 
The polygon incidence data can be seen in Panel A of Figures~\ref{predobsmapsen}--\ref{predobsmapidn}.

We selected Indonesia, Senegal and Madagascar as case examples as they all have abundant subnational surveillance data and country-wide surveys from approximately the same periods.
To minimise temporal effects we selected one year of polygon incidence data and the surrounding five years of prevalence point-survey data for each country.
Within this five year period, we considered malaria unchanging and did not model time explicitely.
For Indonesia we selected polygon incidence data from 2012 that covers 379 administrative units, and prevalence data from 2010 to 2014 that consists of 1,233 survey clusters (i.e. unique locations), representing 230,747 individuals.
For Senegal we selected 2015 for polygon incidence data (41 administrative units) and 2013 to 2017 for prevalence data (804 surveys, 17,037 individuals).
Finally, for Madagascar we selected 2013 for polygon incidence (110 administrative units) and 2011 to 2015 for prevalence data (1,049 surveys, 36,411 individuals).

@@

We assigned $\sigma_v$ a penalised complexity prior \citep{simpson2017penalising} such that $P(\sigma_v > 0.05) = 0.0000001$
This was based on a comparison of the variance of Poisson random variables, with rates given by the number of cases observed, and an independently derived upper and lower bound for the case counts using the approach defined by Cibulskis and colleagues \citep{cibulskis2011worldwide}.
We found that an iid effect with a standard deviation of 0.05 was able to account for the discrepancy between the assumed Poisson error and the independently derived error.

@@

Future models could potentially be improved by using a more flexible approach for addressing the shortcomings in the prevalence-incidence relationship \citep{cameron2015defining}.
This could be by estimating the parameters of the polynomial jointly with the rest of the model.
Informative priors based on the original model could be used to regularise this joint fit both to prevent improbable inferences but also because if the relationship were too flexible, the information from the prevalence data might not contribute to informing the regression parameters and spatial random field.
This is particularly true for model forms such as a spline or a Gaussian process on the relationship between prevalence and incidence.
For the model to handle noisy or biased prevalence point-surveys, the modeller can control the iid random effect on the point-surveys, $w_b$ and the prevalence intercept $\beta_p$. 
Here we have tried to maximise the influence of the prevalence data by setting the prior based on the belief that the random effect should only explain extra-binomial variation that is impossible to derive from the covariates (e.g. based on the differences in prevalence surveys within the same pixel).
Weakening this prior will allow the iid effect to explain more of the prevalence point-survey variation which both reduces the potential statistical power gained by adding the point-surveys but also reduces the effects of biased or noisy estimates.

@@

We used three case studies, limited by the number of coutnries with good aggregated incidence data as well as good prevalence survey data.
Given the small number of case studies it is hard to determine when these methods are likely to be most effective.
However, the greatest benefits here were seen in Madagascar, a country with more intermediate transmission intensities.
In the future, two groups of countries might particularly benefit from the methods presented here.
Firstly countries who have had large prevalence surveys in the past and whose reporting systems are improving, such as Ethiopia, might benefit from these methods.
Secondly, countries that have lots of prevalence surveys and are adjacent to countries with good reporting systems, (e.g. Papua New Guinea and neighbouring Indonesia), might also benefit from models that share information between countries.

>>>>

Global malaria incidence has decreased dramatically over the last 20 years \citep{bhatt2015effect, weiss2019mapping, battle2019mapping}.
This decrease has been accompanied by a strategic shift aiming for elimination in low incidence countries \citep{world2016world, newby2016path}.
Accurate, high-resolution maps of malaria risk are vital in countries in the elimination and pre-elimination phases as they highlight the areas with ongoing \emph{Plasmodium} transmission most in need of interventions \citep{sturrock2016mapping, cohen2017mapping}.
Mapping malaria in low burden countries presents new challenges as traditional mapping of prevalence \citep{gething2011new, bhatt2017improved, gething2012long, bhatt2015effect} using cluster-level surveys and model-based geostatistics are not necessarily effective in these areas \citep{sturrock2016mapping, sturrock2014fine}.
In low burden areas, very large sample sizes are needed before a prevalence survey is informative because so few individuals have detectable parasitaemia that most sample points will have no cases.
In most cases, these large sample sizes are neither logistically nor financially feasible.
However, the availability and quality of routine surveillance data of malaria case counts, typically aggregated by administrative unit polygons, is improving, thus providing an alternative data source for mapping malaria burden \citep{sturrock2016mapping, ohrt2015information, cibulskis2011worldwide}.
Advantageously, the routine surveillance data can be more sensitive than prevalence point-surveys in low transmission areas because the entire public health system is being used to passively monitor disease occurrence continually over a period of time\citep{cibulskis2011worldwide}.

Disaggregation regression methods have been proposed as a way to model malaria burden using polygon-level, routine surveillance records of incidence \citep{sturrock2014fine, wilson2017pointless, law2018variational, taylor2017continuous, li2012log}.
The intent of this research is to assess the utility of extending disaggregation regression modelling approaches to map malaria incidence using both point- and polygon-level response data. 
In particular it focuses on two aspects where improvements in the disaggregation regression may be possible:
(1) for modelling low-burden areas which have better coverage of prevalence surveys than polygon incidence data; and
(2) for building the statistical relationships between the response (i.e. malaria incidence) and geospatial predictor data in heterogeneous landscapes. 
An ancillary benefit of a hybridized approach is that it will simultaneously produce estimates of both prevalence and incidence metrics, which may both be useful for policy makers \citep{cohen2017mapping}.
Models are typically fitted to observations of one metric and then a secondary model is used to convert between prevalence and incidence \emph{post hoc} \citep{battle2019mapping, bhatt2015effect}, thus missing an opportunity to learn the relationship between prevalence and incidence at the same time as fitting the geographic model.

@@

We used two data sources that quantify malaria burden: prevalence point-surveys and polygon incidence data.
Prevalence point-surveys consist of geo-located survey clusters wherein all sampled individuals are tested for malaria and the positive cases as well as the total number of children tested is recorded. 
Polygon incidence data is aggregated to administrative units (e.g. districts or provinces) summarizing data reported from hospitals and health facilities. 
Unlike the point data, polygon-level reports only include numbers of cases and not the numbers of individuals in each administrative unit.  
As such, to determine an incidence rate we rely on gridded population surfaces, summarised to administrative unit boundaries, to provide the denominator. 
The prevalence point-survey data were extracted from the Malaria Atlas Project database \citep{bhatt2015effect, guerra2007assembling, pfeffer2018ma}. 
As the prevalence point-surveys cover different age ranges they were standardised to the 2--10 year-range using a previously published model \citep{smith2007standardizing}. 
As described, the age standardisation model gives the surveys with zero positive cases a small positive prevalence. 
The polygon incidence data were collated from various government reports and adjusted for incompleteness using methods defined by Cibulskis and colleagues \citep{cibulskis2011worldwide, weiss2019mapping}. 
These adjustments account for underreporting of clinical cases due to lack of treatment seeking, missing case reports (from a health facility that reported for 11 months in a year for example), and cases that sought medical attention outside the public health systems \citep{battle2016treatment}. 
Where species specific reports were given, these were used, and in reports that did not distinguish between species of \emph{Plasmodium} the national estimate of the ratio between \emph{P. falciparum} and \emph{Plasmodium vivax} cases was used to estimate numbers of \emph{P. falciparum} cases specifically. 
These adjustments were uniform across each country. 
The polygon incidence data can be seen in Panel A of Figures~\ref{predobsmapsen}--\ref{predobsmapidn}.

We selected Indonesia, Senegal and Madagascar as case examples as they all have abundant subnational surveillance data and country-wide surveys from approximately the same periods.
To minimise temporal effects we selected one year of polygon incidence data and the surrounding five years of prevalence point-survey data for each country.
Within this five year period, we considered malaria unchanging and did not model time explicitely.
For Indonesia we selected polygon incidence data from 2012 that covers 379 administrative units, and prevalence data from 2010 to 2014 that consists of 1,233 survey clusters (i.e. unique locations), representing 230,747 individuals.
For Senegal we selected 2015 for polygon incidence data (41 administrative units) and 2013 to 2017 for prevalence data (804 clusters, 17,037 individuals).
Finally, for Madagascar we selected 2013 for polygon incidence (110 administrative units) and 2011 to 2015 for prevalence data (1,049 clusters, 36,411 individuals).

@@

We assigned $\sigma_v$ a penalised complexity prior \citep{simpson2017penalising} such that $P(\sigma_v > 0.05) = 0.0000001$.
This was based on a comparison of the variance of Poisson random variables, with rates given by the number of cases observed, and an independently derived upper and lower bound for the case counts using the approach defined by Cibulskis and colleagues \citep{cibulskis2011worldwide}.
We found that an iid effect with a standard deviation of 0.05 was able to account for the discrepancy between the assumed Poisson error and the independently derived error.

@@

Future models could potentially be improved by using a more flexible approach for addressing the shortcomings of the prevalence-incidence relationship \citep{cameron2015defining} being used in this context.
This could be by estimating the parameters of the polynomial jointly with the rest of the model.
Informative priors based on the original model could be used to regularise this joint fit both to prevent improbable inferences but also because if the relationship were too flexible, the information from the prevalence data might not contribute to informing the regression parameters and spatial random field.
This is particularly true for model forms such as a spline or a Gaussian process on the relationship between prevalence and incidence.
For the model to handle noisy or biased prevalence point-surveys, the modeller can control the iid random effect on the point-surveys, $w_b$ and the prevalence intercept $\beta_p$. 
Here we have tried to maximise the influence of the prevalence data by setting the prior based on the belief that the random effect should only explain extra-binomial variation that is impossible to derive from the covariates (e.g. based on the differences in prevalence surveys within the same pixel).
Weakening this prior will allow the iid effect to explain more of the prevalence point-survey variation which both reduces the potential statistical power gained by adding the point-surveys but also reduces the effects of biased or noisy estimates.

@@

We used three case studies, limited by the number of countries with good aggregated incidence data as well as good prevalence survey data.
Given the small number of case studies it is hard to determine when these methods are likely to be most effective.
However, the greatest benefits here were seen in Madagascar, a country with more intermediate transmission intensities.
In the future, two groups of countries might particularly benefit from the methods presented here.
Firstly countries who have had large prevalence surveys in the past and whose reporting systems are improving, such as Ethiopia, might benefit from these methods.
Secondly, countries that have lots of prevalence surveys and are adjacent to countries with good reporting systems, (e.g. Papua New Guinea and neighbouring Indonesia), might also benefit from models that share information between countries.
